Constructive Preference Elicitation by Setwise Max-margin Learning

### 2016-01-28 paper/ijcai16.tex
tagli nella prima parte

Preferences \cite{Peintner2008} play an important role in artificial
intelligence, including fields spanning from recommender systems to
automatic planning, from non monotonic reasoning to computational
social choice and algorithmic decision theory.  The task of eliciting
or learning preferences is therefore a crucial one; typically only limited
information about the user's preferences will be available and the
cost of obtaining additional preference information will be high (from
a cognitive and/or computational point of view).  The goal in
preference elicitation is that of gaining enough information about the
user's
preferences %(typically represented by an utility function $u$, unknown to the system)
in order to recommend a decision or a course of action.

The assessment of preferences have received considerable attention in the literature, with early works confined to the operation research and management literature mostly \cite{White1984,JacquetLagreze95}.
%In the past decade a number of researchers have considered elicitation of preferences from an artificial intelligence perspective%; within the machine learning field the term preference learning has been catching up.
%While preferences are often assessed using some ad-hoc rules \cite{},
Recently, in the last decade or so, a number of artificial
intelligence researchers have proposed principled methods that elicit
utility in an adaptive way
\cite{chajewska2000,boutilier2002,Wang2003,boutilier2006,braziunas2007,guo2010real,viappiani2010optimal}. The
aim is that of focusing on learning the ``important'' part of the
utility, allowing to recommend good (or even optimal) decisions with
only partial knowledge of the utility function.

%braziunas-mmr:uai07,

@@

Preference elicitation in configuration problems has been previously
tackled with regret-based elicitation
\cite{boutilier2006,braziunas2007}, where minimax regret is used both
as a robust recommendation criterion and as a technique to drive
elicitation in configuration problems.  The main limitation of the
minimax regret approach is the lack of tolerance with respect to user
inconsistency (the user might state that an item is preferred to
another one, even though the latter has higher utility).
% setwise max-margin formulation to deal with uncertainty in user utility

Indeed, learning a user utility function requires setting a preference
elicitation strategy and dealing with uncertain and possibly
inconsistent user feedback.  Bayesian preference elicitation
approaches deal with both problems by building a probability
distribution on candidate functions (endowed with a response or error
model to be used for inference) and asking queries maximizing
informativeness measures such as expected value of information
(EVOI)~\cite{chajewska2000,guo2010real,viappiani2010optimal}.  These
approaches are however computationally expensive and can not scale to
fully constructive scenarios, as shown in our experimental results.

We take a space decomposition perspective and jointly learn a set of
weight vectors, each representing a candidate utility function,
maximizing diversity between the vectors and consistency with the
available feedback. These two conflicting objectives tend to generate
equally plausible alternative hypotheses for the unknown
utility. 
Our approach to preference elicitation works by combining weight vector
learning with instance generation, so that each iteration of the
algorithm produces two outcomes: a set of weight vectors and a set of
instances, each maximizing its score according to one of the weight
vectors.

We evaluate the effectiveness of our approach by testing our
elicitation method in both synthetic and real-world problems, and
comparing it to state-of-the-art methods.

The paper is structured as follows: we present some background
material in Section \ref{sec:background} and then proceed to our
setwise max-margin method in Section \ref{sec:formulation}; we discuss
experimental results in Section \ref{sec:experiments} and conclude
with final remarks (Sec.~\ref{sec:conclusions}).

%\section{Related Work}

@@

\paragraph{Notation.} We use boldface letters $\vx$ to indicate vectors,
uppercase letters $X$ for matrices, and calligraphic capital letters $\calX$
for sets. We frequently abbreviate the set $\{ x^i \}_{i=1}^n$ as $\{ x^i \}$
whenever the range of the index $i$ is clear from the context, and use $[n]$ as
a shorthand for the set $\{1, \ldots, n\}$. We write $\|\vx\|_1 := \sum_z |x_z|$
to indicate the $\ell_1$ vector norm, $\langle \cdot, \cdot \rangle$ to
denote the usual dot product, and $X'$ for matrix transposition.

\paragraph{Setting.} We assume to have a multi-attribute feature space
$\calX$ of configurations $\vx = (x_1, \ldots, x_m)$ over $m$
features. For the sake of simplicity we focus on binary features only,
i.e. $x_z\in\{0,1\}$ for all $z\in[m]$, assuming a one-hot encoding of
categorical features. This is a common choice for preference
elicitation methods~\cite{guo2010real,viappiani2010optimal}. Support
for linearly dependent continuous features will be discussed later on.

@@

Consistently with previous work~\cite{guo2010real,viappiani2010optimal}, 
we model users by means of utility functions of a given structure; the user's preferences are represented by a latent weight
vector $\vw\in\bbR^m$.
We assume linear utility functions; the utility of a configuration $\vx$ is given by
$\langle \vw, \vx \rangle = \sum_{z=1}^m w_z x_z$
(see \cite{keeney1976} for theoretical conditions justifying additive utility functions). 
In the remainder of the paper we
require all weights to be {\em non-negative} and {\em bounded}: the per-attribute
weights $w_z$ must lie in a (constant but otherwise arbitrary) interval
$[w^\bot_z, w^\top_z]$, with $w^\bot_z \ge 0$. 
Both requirements are quite natural\footnote{Utility values are defined on an
interval scale, therefore the assumption of non negativity is not restrictive as it is always possible to scale the values appropriately (see for instance \cite{Torra2007} and \cite{keeney1976}).}, and enable the translation of our core optimization problem into a
mixed-integer linear problem (as done in Section~\ref{sec:formulation}). 
\andrea{@paolo: sulla giustificazione delle utilita' additive, vogliamo metterla? c'e' anche molto lavoro che contesta il fatto che basti no?}

During learning, the actual weight vector $\vw$ is {\em unknown} to
the learning system, and must be estimated by interacting with the
user. We mostly focus on pairwise comparison queries, which are the
simplest of the comparative queries. These can be extended to choice
sets of more than two options
\cite{viappiani2009,viappiani2010optimal}.  Comparison and choice
queries are common in conjoint analysis
\cite{louviere2000,toubia2004}. We consider three possible outcomes
for a pairwise comparison between two configurations $\vx$ and $\vx'$:
either $\vx$ is preferred to $\vx'$ (written $\vx \succ \vx'$), $\vx'$
is preferred to $\vx$ ($\vx \prec \vx'$), or there is no clear
preference between the two items ($\vx \approx \vx'$). We write
$\calD$ to denote the set of preferences (answers to comparison
queries) elicited from the user. Queries are typically selected so as
to maximize some measure of informativeness of the resulting
responses.  Details on our actual query selection strategy will be
provided in Section~\ref{sec:formulation}.

@@

The goal of our setwise max-margin approach is twofold. First, for any
given set size $k\geq 1$, we want to find a {\em set} of $k$ weight
vectors $\vw^{1}, \ldots, \vw^{k}$, chosen so that all user-provided
preferences are satisfied by the largest possible margin (%
%i.e. all
%weight vectors are consistent with respect to the user responses
%$\calD$, 
modulo inconsistencies) and so that they are maximally
diverse.  Second, we want to construct a {\em set} of $k$
configurations $\vx^{1}, \ldots, \vx^{k}$, so that each configuration
$\vx^{i}$ is the ``best'' possible option when evaluated according to
the corresponding $\vw^{i}$ and configurations are maximally diverse
among each other. These options will be later employed to formulate
the user queries.

@@

The two core constraints are Equations~(\ref{eq:wyconstr}) and~(\ref{eq:wxconstr}),
which are taken directly from the discussion above. The former enforces the
correct ranking of the observed user preferences, while the latter ensures that
the generated configurations are diverse in terms of the weight vectors they
maximize. Equation~(\ref{eq:wbounds}), where the inequality is intended to be component-wise, and Eq.~(\ref{eq:xbounds})
ensure that the weights and configurations are in fact feasible, while the
remaining two guarantee the non-negativity of the slack and margin variables.
Since we require $\vw^\bot \ge (0,\ldots,0)$, Eq.~(\ref{eq:wbounds}) also enforces the weights to be non-negative.

Note that we are choosing the configurations $\{ \vx^i \}$ and the weight vectors $\{
\vw^i \}$ {\em simultaneously}.
Look at Figure~\ref{fig:setmargin}, where, for simplicity, we need to choose a pair ($k=2$). 
The constraint \ref{eq:wxconstr} is represented by a red line, that partitions the space of
feasible utility weights in two parts (in general, there will be $k$ subregions).
Since we maximize the margin $\mu$, the optimizer
will prefer a set of configurations $\{ \vx^{1}, \vx^{2} \}$ that partitions the weight space
in an even way.
In each subregion, we have corresponding utility weight vector $\vw^{i}$
lying ``close'' to the centre of each subregion. 
Note that if we present $\vx^{1}$ and $\vx^{2}$ to the user, and the user states that he prefers
the former to the latter, the feasible region for the weights will then become the part of the polytope to the left of the red line; moreover $\vw^{1}$ will be vector maximizing the margin in the classic sense.
Therefore, setwise maxmargin can be thought as a ``look-ahead'' model with respect to a standard ($k=1$) max-margin learner a posteriori.
%Therefore, setwise maxmargin can be thought as the set that, given any possible preference about the items, ensure that the reduction of utility uncertainty is the greatest
%\paolo{ho fatto un tentativo di migliorare l'intuizione, ma e' ancora possibile migliorare...}
\andrea{cosa intendi per a posteriori?}

\begin{figure}[t]
    \begin{center}
        \includegraphics[width=10em]{figures/setmargin}
    \end{center}
    \caption{\label{fig:setmargin} The intuition behind setwise max-margin; the red points are utility vectors $\vw^{1}$ $\vw^{2}$, the red line corresponds to the constraint $\langle \vw, \vx^{1} - \vx^{2} \rangle \ge 0$.}
\end{figure}

@@

\paragraph{Set-wise max-margin.} The full {\sc SetMargin} algorithm
follows the usual preference elicitation loop. Starting from an
initially empty set of user responses $\calD$, it repeatedly solves
the MILP problem above using the set $\calD$ to enforce ranking
constraints on the weight vectors $\{\vw^i\}$. The generated
configurations $\{\vx^i\}$, which are chosen to be as good as possible
with respect to the estimated user preferences, and as diverse as
possible, are then employed to formulate a set of user queries. The
new replies are added to $\calD$ and the whole procedure is
repeated. The algorithm terminates after a fixed number of iterations,
or once another suitable condition is met.  Termination might be left
to the user to decide (see for example \cite{Reilly2007});
alternatively we can terminate when the difference between the utility
weights is small enough.

@@

We presented a max-margin approach for efficient preference
elicitation in large configuration spaces. Our approach relies on an
extension of max-margin learning to sets of hyperplanes, and is
effective in the generation of a diverse set of items that can be used
to ask informative queries to the user.  The main advantages of our
elicitation method are 1) its ability to elicit preferences and
provide recommendations in large configuration problems 2) its
robustness with respect to erroneous feedback and 3) its ability to
encourage sparse utility functions. Experimental comparisons against
state-of-the-art Bayesian preference elicitation strategies confirm
its substantial computational advantages. For sparse utility
functions, the method compares favourably also in terms of the number
of queries asked to the user.

>>>>

Preferences \cite{Peintner2008} play an important role in a variety of artificial
intelligence applications and the task of eliciting or learning preferences is a crucial one; 
typically only limited
information about the user's preferences will be available and the
cost (cognitive or computational) obtaining additional preference information will be high.  
%The goal in
%preference elicitation is that of gaining enough information about the
%user's
%preferences %(typically represented by an utility function $u$, unknown to the system)
%in order to recommend a decision or a course of action.

The automated assessment of preferences has received considerable attention,
starting with pioneering works \cite{White1984,JacquetLagreze95} in the OR community.
%In the past decade a number of researchers have considered elicitation of preferences from an artificial intelligence perspective%; within the machine learning field the term preference learning has been catching up.
%While preferences are often assessed using some ad-hoc rules \cite{},
More recently, a number of AI researchers have proposed interactive methods that elicit
preferences in an adaptive way
\cite{chajewska2000,boutilier2002,Wang2003,boutilier2006,guo2010real,viappiani2010optimal},
observing that, by asking informative questions, it is often possible to make near-optimal decisions with only partial preference information.

%focusing on learning the ``important'' part of the
%preferences, 
%braziunas-mmr:uai07,

@@

Preference elicitation in configuration problems has been previously
tackled with regret-based elicitation
\cite{boutilier2006,braziunas2007}, where minimax regret is used both
as a robust recommendation criterion and as a technique to drive
elicitation.  The main limitation of their
approach is the lack of tolerance with respect to user
inconsistency. %(the user might state that an item is preferred to
%another one, even though the latter has higher utility).
% setwise max-margin formulation to deal with uncertainty in user utility
Indeed, learning a user utility function requires 
%elicitation strategy and 
to deal with uncertain and possibly inconsistent user feedback.

Bayesian preference elicitation
approaches deal with this problem by 
building a probability
distribution on candidate functions (endowed with a response or error
model to be used for inference) and asking queries maximizing
informativeness measures such as {\em expected value of information
(EVOI)}~\cite{chajewska2000,guo2010real,viappiani2010optimal}.  
These
approaches are however computationally expensive and can not scale to
fully constructive scenarios, as shown in our experimental results.

We take a space decomposition perspective and jointly learn a set of
weight vectors, each representing a candidate utility function,
maximizing diversity between the vectors and consistency with the
available feedback. These two conflicting objectives tend to generate
equally plausible alternative hypotheses for the unknown
utility. 
Our approach to elicitation works by combining weight vector
learning with instance generation, so that each iteration of the
algorithm produces two outcomes: a set of weight vectors and a set of
instances, each maximizing its score according to one of the weight
vectors. 
We evaluate the effectiveness of our approach by testing our
elicitation method in both synthetic and real-world problems, and
comparing it to state-of-the-art methods.

% The paper is structured as follows: we present some background
% material in Section \ref{sec:background} and then proceed to our
% setwise max-margin method in Section \ref{sec:formulation}; we discuss
% experimental results in Section \ref{sec:experiments} and conclude
% with final remarks (Sec.~\ref{sec:conclusions}).

@@

%\paragraph{Notation.}
We first introduce some notation.
We use boldface letters $\vx$ to indicate vectors,
uppercase letters $X$ for matrices, and calligraphic capital letters $\calX$
for sets. We %frequently 
abbreviate the set $\{ x^i \}_{i=1}^n$ as $\{ x^i \}$
whenever the range of the index $i$ is clear from the context, and use $[n]$ as
a shorthand for %the set 
$\{1, \ldots, n\}$. We write $\|\vx\|_1 := \sum_z |x_z|$
to indicate the $\ell_1$ vector norm, $\langle \cdot, \cdot \rangle$ for the usual dot product,  $X'$ for matrix transposition.

%\paragraph{Setting.} 
We assume to have a multi-attribute feature space
$\calX$ of configurations $\vx = (x_1, \ldots, x_m)$ over $m$
features. For the sake of simplicity we focus on binary features only,
i.e. $x_z\in\{0,1\}$ for all $z\in[m]$, assuming a one-hot encoding of
categorical features. This is a common choice for preference
elicitation methods~\cite{guo2010real,viappiani2010optimal}. Support
for linearly dependent continuous features will be discussed later on.

@@

Consistently with previous work~\cite{guo2010real,viappiani2010optimal}, 
we model users with additive utility functions \cite{keeney1976}; 
the user's preferences are represented by a  weight
vector $\vw\in\bbR^m$ and
%We assume linear utility functions \cite{keeney1976}; 
the utility of a configuration $\vx$ is given by
$\langle \vw, \vx \rangle = \sum_{z=1}^m w_z x_z$. 
In the remainder of the paper we
require all weights to be {\em non-negative} and {\em bounded}: the per-attribute
weights $w_z$ must lie in a (constant but otherwise arbitrary) interval
$[w^\bot_z, w^\top_z]$, with $w^\bot_z \ge 0$. 
Both requirements are quite natural\footnote{Utility values are defined on an
interval scale, therefore the assumption of non negativity is not restrictive as it is always possible to scale the values appropriately (see for instance \cite{Torra2007} and \cite{keeney1976}).}, and enable the translation of our core optimization problem into a
mixed-integer linear problem (as done in Section~\ref{sec:formulation}).

During learning, the actual weight vector $\vw$ is {\em unknown} to
the learning system, and must be estimated by interacting with the
user. We mostly focus on pairwise comparison queries, which are the
simplest of the comparative queries. These can be extended to choice
sets of more than two options
\cite{viappiani2009,viappiani2010optimal} and are common in conjoint analysis
\cite{louviere2000,toubia2004}. %We consider three possible outcomes
For a pairwise comparison between two configurations $\vx$ and $\vx'$:
either $\vx$ is preferred to $\vx'$ (written $\vx \succ \vx'$), $\vx'$
is preferred to $\vx$ ($\vx \prec \vx'$), or there is no clear
preference between the two items ($\vx \approx \vx'$). We write
$\calD$ to denote the set of preferences (answers to comparison
queries) elicited from the user.

In the next Section we describe how informative queries can be generated
using our setwise maxmargin learning.
% Queries are typically selected so as
% to maximize some measure of informativeness of the resulting
% responses.  In the next Section we describe how Details on our actual query selection strategy will be
% provided in Section~\ref{sec:formulation}.

@@

The goal of our setwise max-margin approach is twofold. First, for any
given set size $k\geq 1$, we want to find a {\em set} of $k$ weight
vectors $\vw^{1}, \ldots, \vw^{k}$, chosen so that all user-provided
preferences are satisfied by the largest possible margin (%
%i.e. all
%weight vectors are consistent with respect to the user responses
%$\calD$, 
modulo inconsistencies) and so that they are maximally
diverse.  Second, we want to construct a {\em set} of $k$
configurations $\vx^{1}, \ldots, \vx^{k}$, so that each configuration
$\vx^{i}$ is the ``best'' possible option when evaluated according to
the corresponding $\vw^{i}$ and configurations are maximally diverse
among each other. These options will be later used to formulate
queries.

@@

The two core constraints are~(\ref{eq:wyconstr}) and~(\ref{eq:wxconstr}),
which are taken directly from the discussion above. The former enforces the
correct ranking of the observed user preferences, while the latter ensures that
the generated configurations are diverse in terms of the weight vectors they
maximize. Constraint~(\ref{eq:wbounds}), intended to be component-wise, and~(\ref{eq:xbounds})
ensure that the weights and configurations are feasible, while the
remaining two guarantee the non-negativity of the slack and of the margin.
Since we require $\vw^\bot \ge (0,\ldots,0)$, Eq.~(\ref{eq:wbounds}) also enforces the weights to be non-negative.

Note that we are choosing the configurations $\{ \vx^i \}$ and the weight vectors $\{
\vw^i \}$ {\em simultaneously}.
Look at Figure~\ref{fig:setmargin}, where, for simplicity, we need to choose a pair ($k=2$). 
The constraint \ref{eq:wxconstr} is represented by a red line, that partitions the space of
feasible utility weights in two parts (in general, there will be $k$ subregions).
Since we maximize the margin $\mu$, the optimizer
will prefer a set of configurations $\{ \vx^{1}, \vx^{2} \}$ that partitions the weight space
in an even way.
In each subregion, we have corresponding utility weight vector $\vw^{i}$
lying ``close'' to the centre of each subregion. 
If we then present $\vx^{1}$ and $\vx^{2}$ to the user, and the user states that he prefers
the former to the latter, the feasible region for the weights will then become the part of the polytope to the left of the red line; moreover $\vw^{1}$ will maximize the margin in the classic sense ($k=1$).
%Therefore, setwise maxmargin can be thought as a ``look-ahead'' model with respect to a standard ($k=1$) max-margin learner.
%Therefore, setwise maxmargin can be thought as the set that, given any possible preference about the items, ensure that the reduction of utility uncertainty is the greatest
%\paolo{ho fatto un tentativo di migliorare l'intuizione, ma e' ancora possibile migliorare...}

\begin{figure}[t]
    \begin{center}
        \includegraphics[width=10em]{figures/setmargin}
    \end{center}
    \caption{\label{fig:setmargin} Optimization of setwise max-margin; the black lines corresponds to preference constraints $\calD$, the red points are utility vectors $\vw^{1}$ $\vw^{2}$, the red line corresponds to the constraint $\langle \vw, \vx^{1} - \vx^{2} \rangle \ge 0$.}
\end{figure}

@@

\paragraph{Set-wise max-margin.} The full {\sc SetMargin} algorithm
follows the usual preference elicitation loop. Starting from an
initially empty set of user responses $\calD$, it repeatedly solves
the MILP problem above using $\calD$ to enforce ranking
constraints on the weight vectors $\{\vw^i\}$. The generated
configurations $\{\vx^i\}$, which are chosen to be as good as possible
with respect to the estimated user preferences, and as diverse as
possible, are then employed to formulate a set of user queries. The
new replies are added to $\calD$ and the whole procedure is
repeated. Termination can be after a fixed number of iterations,
when the difference between utility vectors is very small, or might be left
to the user to decide (e.g. \cite{Reilly2007}).

@@

We presented a max-margin approach for efficient preference
elicitation in large configuration spaces. Our approach relies on an
extension of max-margin learning to sets, and is
effective in the generation of a diverse set of configurations that can be used
to ask informative preference queries.  The main advantages of this
elicitation method are 1) its ability to elicit preferences and
provide recommendations in large configuration problems 2)
robustness with respect to erroneous feedback and 3)  ability to
encourage sparse utility functions. 
Experimental comparisons against
state-of-the-art Bayesian preference elicitation strategies confirm
its substantial computational advantages. For sparse utility
functions, the method compares favorably also in terms of utility loss.
%the number
%of queries asked to the user.
